Fast Random Integer Generation in an Interval

### 2024-01-16 Paper/main.tex
finalize report

\category{} 
\relatedversion{} 
\supplement{Sources of implementations and experiments can be accessed at \url{https://github.com/lukasgeis/FRIGiaI-Review}}
\acknowledgements{}
\nolinenumbers

@@

%%%%% Abstract
\begin{abstract}
    In randomized algorithms and other probabilistic methods we often need to generate uniform random integers in a given interval.
    This is essential for algorithms like the Fisher-Yates random shuffle or simply drawing an element from an array uniformly at random.
    Pseudo-random numbers however are often generated as words with a fixed number of bits and thus need to be mapped to the desired interval without introducing bias.
    Common methods often include division operations that are relatively expensive on most processors.
    Therefore, Lemire introduced an algorithm that can sample uniform random integers in a given range $[0,n)$ with zero to one division operations at most~\cite{Lemire}.

@@

%%%%% Introduction
\section{Introduction}
\subsection{Motivation}
The study of randomized algorithms goes as far back as the 1910s~\cite{MathComp} and is now scientifically pursued in almost all parts of computer science.
Not only in theory, but also in practice, randomized algorithms gain even more importance today to not only solve specific algorithmic problems but also to create large, scalable, and unbiased data sets that can be used in a variety of fields.
\\
The core of such algorithms is often the pseudo-random number generator (PRNG).
There are many efficient PRNGs such as XorShift~\cite{XS}, MersenneTwister~\cite{MT}, linear congruential generators~\cite{LinConGen, MINSTD, LM}, and younger generators such as the PCG-family~\cite{PCG} and ChaCha~\cite{Cha}.
These generators produce $32$-bit or $64$-bit words which we can interpret as unsigned integers in $[0,2^{32})$ and $[0,2^{64})$ respectively.
However, for most applications we do not want to have such a large range of possible random values but instead want to confine the range to a given interval $[a,b]$.
This can not only be used to uniformly sample an element from an array but also for more complex algorithms
\begin{itemize}
    \item The Fisher-Yates random shuffle~\cite{FY} (see \hyperref[alg:fy]{Algorithm 1}) uniformly permutes an array of $n$ elements in time $\Theta(n)$ if we can sample in constant time. 
    We have to draw a random integer in an interval for every shuffled element.
    \item A random graph model describes a distribution over a (parametrized) family of graphs. 
    There exist an uncountable number of such models and almost all of them require some sort of integer sampling in an interval. 
    For example, the $\mathcal{G}(n,m)$ model~\cite{GNM} uniformly samples a (directed) graph with $n$ nodes and $m$ edges. 
    A possible algorithm for this problem is Reservoir-Sampling (see \hyperref[alg:gnm]{Algorithm 2})~\cite{Reservoir} which is a general technique that can sample $k$ elements without replacement from an input stream. 
    Another model is the BA model~\cite{BABook,BA} which iteratively builds the graph by adding nodes with edges to previously added nodes. %(see \hyperref[alg:ba]{Algorithm 3}). 
    Both models require sampling of an integer in an interval for each of its edges (or more).
    \item An Alias-Table~\cite{Alias} can be used to simulate static discrete distributions in practice. 
    After creating and initializing the table, we can sample in time $\Theta(1)$ by drawing a uniform random row and offset in each round. 
    Hence, we need to sample an integer uniformly at random in $[0,n)$ where $n$ is the number of rows in the table.
\end{itemize}

@@

\subsection{Preliminaries}\label{sec:1.2}
We denote by $x \div y \coloneqq \lfloor\nicefrac{x}{y}\rfloor$ the integer division operation, and the remainder operation by $x \Mod y \coloneqq x - (x \div y)y$. 
Using a Bit-\textsc{RightShift}, denoted by $\gg$, we can efficiently implement divisions by a power of two: $x \div 2^W \coloneqq x \gg W$.
Similarly, we denote a Bit-\textsc{LeftShift} by $\ll$.
Additionally, with a bitwise \textsc{And}-operation, denoted by $\&$, we can efficiently compute the remainder of a division by a power of two: $x \Mod 2^W \coloneqq x \And (2^W - 1)$.
In practice, PRNGs generate uniform random numbers in an interval $[0,2^W)$ where $W \in \{32,64\}$, namely a uniform random $W$-bit number.

@@

\subsubsection{Modulo Reduction}\label{sec:2.1.1}
The most common and under-egineered algorithm is the modulo reduction, namely
\begin{center}
    $\texttt{rand()} \Mod n$
\end{center}
Clearly, this method returns a random integer in the interval $[0,n)$.
However, as seen in \hyperref[eq:1]{equation (1)}, if $n$ does not perfectly divide $2^W$, we have a \emph{leftover} interval at the end that introduces bias as not all integers in $[0,n)$ are a part of this interval.
Thus, not all integers in $[0,n)$ are equally likely and this reduction does not produce uniform numbers.

@@

Again, this method clearly generates a random integer in $[0,n)$ because $y$ is less than $1$ and we multiply with $n$ before rounding down.
However, in the commonly implemented floating-point standard (IEEE 754) with $32$ bits, we have a mantisse of $24$ bits and hence can at best represent all values in $[0,2^{24})$ divided by $2^{24}$.
Thus, we can not represent all numbers in $[0,2^{32})$ divided by $2^{32}$ and have less precision than a $32$-bit number.
Hence can not generate all numbers in $[0,n)$ if $n > 2^{24}$.
A solution would be to instead use double precision floating-point numbers that have a mantisse of $53$ bits and thus can represent all values in $[0,2^{53})$ divided by $2^{53}$.
But this again is not enough if $W = 64$ where we require a mantisse of at least $64$ bits which is not accessible on all systems (long double).
Furthermore, the conversion also comes at a price: the corresponding instructions have a latency of at least $6$ cycles on Skylake processors~\cite{Instructions}.

@@

\subsubsection{OpenBSD}\label{sec:2.2.1}
We somehow want to get rid of the bias generated by the last interval in \hyperref[eq:1]{equation (1)}.
The key idea is that this interval does not have to be on the right.
Instead of rejecting all values generated by a random $x \geq 2^W - (2^W \Mod n)$, we could also reject all values generated by a random $x < 2^W \Mod n$ for uniformity since the interval $[2^W \Mod n, 2^W)$ is continuous and has size $2^W - (2^W \Mod n)$ which is divisible by $n$.
Due to \hyperref[lemma:1]{Lemma 1}, for every $r \in [0,n)$, we have exactly $(b - a) \div n$ numbers in $[2^W \Mod n,2^W)$ with remainder $r$, thus having a uniform distribution over $[0,n)$ when computing the remainder.
This yields \hyperref[alg:openbsd]{Algorithm 3} which stems from the C standard library in OpenBSD and macOS (\texttt{arc4random\_uniform}), and has been adopted by the Go language (\texttt{Int63n, Int31n}) with slight implementation differences~\cite{GoLang} as well as the GNU C++ standard library~\cite{GnuCpp}.

@@

\noindent Note that the remainder operation in line $3$ and $9$ can be computed by a bitwise \textsc{And} operation and the division operation in line $10$ by a Right-\textsc{BitShift} as mentioned in \hyperref[sec:1.2]{Section 1}
Thus, the only \emph{real} division operation is the remainder operation in line $5$.

>>>>

\category{} 
\relatedversion{} 
\supplement{Sources can be accessed at \url{https://github.com/lukasgeis/FRIGiaI-Review}}
\acknowledgements{}
\nolinenumbers

@@

%%%%% Abstract
\begin{abstract}
    In randomized algorithms and other probabilistic methods, we often need to generate uniform random integers in a given interval.
    This is essential for algorithms like the Fisher-Yates random shuffle or simply drawing an element from an array uniformly at random.
    Pseudo-random numbers however are often generated as words with a fixed number of bits and thus need to be mapped to the desired interval without introducing bias.
    Common methods often include division operations that are relatively expensive on most processors.
    Therefore, Lemire introduced an algorithm that can sample uniform random integers in a given range $[0,n)$ with zero to one division operations at most~\cite{Lemire}.

@@

%%%%% Introduction
\section{Introduction}
\subsection{Motivation}
The study of randomized algorithms goes as far back as the 1910s~\cite{MathComp} and is now scientifically pursued in almost all parts of computer science.
Not only in theory, but also in practice, randomized algorithms gain even more importance today to not only solve specific algorithmic problems but also to create large, scalable, and unbiased data sets that can be used in a variety of fields.
\\
The core of such algorithms is often the pseudo-random number generator (PRNG).
There are many efficient PRNGs such as XorShift~\cite{XS}, MersenneTwister~\cite{MT}, linear congruential generators~\cite{LinConGen, MINSTD, LM}, and younger generators such as the PCG-family~\cite{PCG} and ChaCha~\cite{Cha}.
These generators produce $32$-bit or $64$-bit words which we can interpret as unsigned integers in $[0,2^{32})$ and $[0,2^{64})$ respectively.
However, for most applications we do not want to have such a large range of possible random values but instead want to confine the range to a given interval $[a,b]$.
This can not only be used to uniformly sample an element from an array but also for more complex algorithms:
\begin{itemize}
    \item The Fisher-Yates random shuffle (see \hyperref[alg:fy]{Algorithm 1})~\cite{FY} uniformly permutes an array of $n$ elements in time $\Theta(n)$ if we can sample in constant time. 
    We have to draw a random integer in an interval for every shuffled element.
    \item A random graph model describes a distribution over a (parametrized) family of graphs. 
    There exist an uncountable number of such models and almost all of them require some sort of integer sampling in an interval. 
    For example, the $\mathcal{G}(n,m)$ model~\cite{GNM} uniformly samples a (directed) graph with $n$ nodes and $m$ edges. 
    A possible algorithm for this problem is Reservoir-Sampling (see \hyperref[alg:gnm]{Algorithm 2})~\cite{Reservoir} which is a general technique that can sample $k$ elements without replacement from an input stream. 
    Another model is the BA model~\cite{BABook,BA} which iteratively builds the graph by adding nodes with edges to previously added nodes. %(see \hyperref[alg:ba]{Algorithm 3}). 
    Both models require sampling of an integer in an interval for each of its edges (or more).
    \item An Alias-Table~\cite{Alias} can be used to simulate static discrete distributions in practice. 
    After creating and initializing the table, we can sample in time $\Theta(1)$ by drawing a uniform random row and offset in each round. 
    Hence, we need to sample an integer uniformly at random in $[0,n)$ where $n$ is the number of rows in the table.
\end{itemize}

@@

\subsection{Preliminaries}\label{sec:1.2}
We denote by $x \div y \coloneqq \lfloor\nicefrac{x}{y}\rfloor$ the integer division operation, and the remainder operation by $x \Mod y \coloneqq x - (x \div y)y$. 
Using a Bit-\textsc{RightShift}, denoted by $\gg$, we can efficiently implement divisions by a power of two: $x \div 2^W \coloneqq x \gg W$.
Similarly, we denote a Bit-\textsc{LeftShift} by $\ll$ which can be used for efficient multiplication by a power of two.
Additionally, with a bitwise \textsc{And}-operation, denoted by $\&$, we can efficiently compute the remainder of a division by a power of two: $x \Mod 2^W \coloneqq x \And (2^W - 1)$.
In practice, PRNGs generate uniform random numbers in an interval $[0,2^W)$ where $W \in \{32,64\}$, namely a uniform random $W$-bit number.

@@

\subsubsection{Modulo Reduction}\label{sec:2.1.1}
The most common and under-egineered algorithm is the modulo reduction, namely
\begin{center}
    $\texttt{rand()} \Mod n$
\end{center}
Clearly, this method returns a random integer in the interval $[0,n)$.
However, as seen in \hyperref[eq:1]{equation (1)}, if $n$ does not divide $2^W$, we have a \emph{leftover} interval at the end that introduces bias as not all integers in $[0,n)$ are a part of this interval.
Thus, not all integers in $[0,n)$ are equally likely and this reduction does not produce uniform numbers.

@@

Again, this method clearly generates a random integer in $[0,n)$ because $y$ is at most $1.0$ and we multiply with $n$ before rounding down (floating-point-errors are negligible because they get corrected in (3) by rounding down).
However, in the commonly implemented floating-point standard (IEEE 754) with $32$ bits, we have a mantisse of $24$ bits and hence can at best represent all values in $[0,2^{24})$ divided by $2^{24}$.
Thus, we can not represent all numbers in $[0,2^{32})$ divided by $2^{32}$ and have less precision than a $32$-bit number.
Hence can not generate all numbers in $[0,n)$ if $n > 2^{24}$.
A solution would be to instead use double precision floating-point numbers that have a mantisse of $53$ bits and thus can represent all values in $[0,2^{53})$ divided by $2^{53}$.
But this again is not enough if $W = 64$ where we require a mantisse of at least $64$ bits which is not accessible on all systems (long double).
Furthermore, the conversion also comes at a price: the corresponding instructions have a latency of at least $6$ cycles on Skylake processors~\cite{Instructions}.

@@

\subsubsection{OpenBSD}\label{sec:2.2.1}
We somehow want to get rid of the bias generated by the last interval in \hyperref[eq:1]{equation (1)}.
The key idea is that this interval does not have to be on the right.
Instead of rejecting all values generated by a random $x \geq 2^W - (2^W \Mod n)$, we could also reject all values generated by a random $x < 2^W \Mod n$ for uniformity since the interval $[2^W \Mod n, 2^W)$ is continuous and has size $2^W - (2^W \Mod n)$ which is divisible by $n$.
Due to \hyperref[lemma:1]{Lemma 1}, for every $r \in [0,n)$, we have exactly $(2^W - 2^W \Mod n) \div n$ numbers in $[2^W \Mod n,2^W)$ with remainder $r$, thus having a uniform distribution over $[0,n)$ when computing the remainder.
This yields \hyperref[alg:openbsd]{Algorithm 3} which stems from the C standard library in OpenBSD and macOS (\texttt{arc4random\_uniform}), and has been adopted by the Go language (\texttt{Int63n, Int31n}) with slight implementation differences~\cite{GoLang} as well as the GNU C++ standard library~\cite{GnuCpp}.

@@

\newpage % In the final version, this paragraph started on the previous page before the algorithm which could be confusing
\noindent Note that the remainder operation in line $3$ and $9$ can be computed by a bitwise \textsc{And} operation and the division operation in line $10$ by a Right-\textsc{BitShift} as mentioned in \hyperref[sec:1.2]{Section 1}.
Thus, the only \emph{real} division operation is the remainder operation in line $5$.
